---
title: "Day 2 exercise"
author: "James Thorson"
date: "2025-05-20"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Arrow-and-lag interface for simultaneous and lagged effects

Agenda section E:  Exercise
Goal:

1.	Understand arrow-and-lag notation

2.	Fit simultaneous and lagged effects in a time-series model

Structure:

1.	Students are to self-organize into groups of 2-3 people, and work in a group on step-1 for 20 minutes.  

2.  For Step-0, we will walk through code as a group;

3.  For later steps, the instructor will then share code for that step on github, and groups will continue with step-2 for 20 minutes, then repeating for step-3.  



## Step 0:  Try fitting a linear model (20 minutes)

Try simulating data from a linear model:

$$
\begin{gathered}
x_i \sim \mathrm{Normal}(2, \sigma_X^2 ) \\
y_i \sim \mathrm{Normal}(1 + 0.5, \sigma_Y^2)
\end{gathered}
$$

Where:

*	$i \in {1,2,…,100}$, $x$ and $y$ are vectors with length 100

*	$\sigma_X^2=1$ is the variance in the predictor variable $x$ and $\sigma_Y^2=1$ is the residual variance in the response variable $y$

*	$E(x_i )=2$ and $E(y_i)=1+0.5x_i$

*	$x_i \sim \mathrm{Normal}(2,\sigma_X^2 )$ indicates that $x_i$ is simulated from a normal distribution, which can be done in R using `rnorm( n = 100, mean = 2, sd = 1)`


```{r simulate_linear_model}
# Simulate x and y
x = rnorm( n = 100, mean = 2, sd = 1 )
y = rnorm( n = 100, mean = 1 + 0.5*x, sd = 1 )
```

Then try fitting in `dsem`:
```{r linear_model}
#install.packages("dsem")
library(dsem)

# Define arrow-and-lag notation
time_term = "
# predictor -> response, lag, parameter_name, start_value
x -> y, 0, slope, 1
x <-> x, 0, sd_x, 1
y <-> y, 0, sd_y, 1
"

# format data into ts object
tsdata = ts(data.frame(x=x, y=y))

# fit in dsem
dsem_fit = dsem(
  tsdata = tsdata, 
  sem = time_term,
  control = dsem_control(quiet = TRUE)
)

# Summarize dsem output
summary(dsem_fit)
```

You can then compare this with a linear model
```{r compare}
# linear model
lm_fit = lm( y ~ 1 + x )
summary(lm_fit)
```

You can also plot output to see the graphical model
```{r plot}
# plot dsem
plot( dsem_fit, 
      edge_label = "value",
      style = "igraph" )
```

## Step 1:  Simulate data (20 minutes)
Please write an R script that simulates abundance at age $n_{a,t}$ for ages $a \in {1,2,…,10}$ and years $t \in {1,2,…,40}$ with the following simplified dynamics:

$$
n_{a,t}=\begin{cases}
  \mu_R \times e^{\sigma_1 \epsilon_{a,t}-ma} & \mathrm{if} \ a=1 \ \& \ t=1 \  \\    
  n_{a-1,t-1} \times e^{\sigma_2 \epsilon_{a,t}-m} & \mathrm{otherwise}
\end{cases}
$$
Where process errors $\epsilon_{a,t} \sim \mathrm{Normal}(0,\sigma^2)$, median age-0 recruitment $\mu_R=10^9$ [individuals], recruitment variation $\sigma_1=0.6$ [dimensionless], demographic variation $\sigma_2=0.1$ [dimensionless], and natural mortality $m=0.4$ [per year].

Now imagine that we can only sample abundance-at-age from a sampling gear with logistic selectivity:
$$
s_a = \frac{1}{1+e^{-\theta_2 (a-\theta_1 ) } }
$$
Where age at 50% selection $\theta_1=3$ [years] and the logit-slope $\theta_2=1$ [year^(-1)].  We then observe:

$$
n_{a,t}^* = n_{a,t} s_a
$$

## Step 2:  Fit model using dsem (20 minutes)
Write out the arrow-and-lag notation whereby $n_{a,t}$ is caused by abundance $n_{a-1,t-1}$ the preceding age and year.  This will presumably involve several one-headed arrows:

```{r example, eval = FALSE}
X -> Y, 1, parameter
```

Where `X` is the predictor, `Y` the response, `1` represents a time-lag, and `parameter` is a parameter name (you can use the same name multiple times to force a single parameter value to be estimated and shared across lines)

Next, modify this arrow-and-lag notation to represent the assumption that exogenous variation at age1 (representing recruitment deviations) will be different than subsequent ages (representing demographic variation).  This will presumably involve several two-headed arrows:

```{r example_2, eval = FALSE}
X <-> X, 0, parameter
```
where these always have a lag of zero.  

## Step 3:  Retrospective skill testing (20 minutes)
Finally, conduct a “leave-future-out” crossvalidation by dropping the final five years of simulated data.  To do so, modify `tsdata` by replacing values (measurements) with NAs (representing missing data).  This will ensure that the model still predicts those missing values.  
